---
title: "Lab 6: Childcare Costs in California"
author: "Catalina Dorado"
date: "April 14, 2025"
format:
  html: default
  gfm: default
execute:
  message: false
  warning: false
  embed-resources: true
  code-tools: true
  echo: true
---

## The Data

In this lab we're going look at the median weekly cost of childcare in
California. A detailed description of the data can be found [here](https://github.com/rfordatascience/tidytuesday/blob/master/data/2023/2023-05-09/readme.md).

The data come to us from [TidyTuesday](https://github.com/rfordatascience/tidytuesday).

**0. Load the appropriate libraries and the data.**

```{r}
#| label: setup
library(tidyverse)
library(dplyr)
```

```{r, message = FALSE}
#| label: load-data

childcare_costs <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-05-09/childcare_costs.csv')

counties <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-05-09/counties.csv')

childcare_costs
counties
```

**1. Briefly describe the dataset (~ 4 sentences). What information does it contain?**

The first dataset, `childcare_costs`, contains numerical values representing various explanatory variables involved in childcare costs. Every variable has a data type as doubles. Each row is identified by their `county_fips_code`, which uniquely identifies the county in a state,  and `study_year`, which is the year the data collection is representative of. The unique identity of each county in a state can be referred to the `counties` dataset, where it contains the FIPS code of the state to which the county belongs to.


## California Childcare Costs

Let's start by focusing only on California. 

**2. Create a `ca_childcare` dataset of childcare costs in California, containing (1) county information and (2) just the year and childcare cost variable information from the `childcare_costs` dataset.** 

*Hint: There are 58 counties in CA and 11 years in the dataset. Therefore, your new dataset should have 53 x 11 = 638 observations. The final data set should have study year, median household income expressed in 2018 dollars, all the variables associated with full-time median price charged for Center-based Care, and California county names*

```{r}
#| label: ca-childcare-costs

childcare <- childcare_costs %>%
  mutate(county_fips_code = as.character(county_fips_code))

ca_counties <- counties %>%
  mutate(county_fips_code = as.character(county_fips_code)) 

ca_childcare <- childcare %>%
  left_join(ca_counties, by = "county_fips_code") %>% 
  filter(state_name == "California")

childcare
ca_counties
ca_childcare
```

**3. Using a function from the `forcats` package, complete the code below to create a new variable where each county is categorized into one of the [10 Census regions](https://census.ca.gov/regions/) in California. Use the Region description (from the plot), not the Region number. An example region has been started for you.**  

*Hint: This is probably a good place to use ChatGPT to reduce on tedious work. But you do need to know how to prompt ChatGPT to make it useful!*

```{r}
#| label: recoding-county-to-census-regions
# Initial code eliminates the word “County” from each of the county names in the `ca_childcare` dataset. Keep and pipe into the rest of your data manipulations.

ca_childcare <- ca_childcare |> 
  mutate(county_name = str_remove(county_name, " County")) |> 
  mutate(region = fct_collapse(county_name, 
                               'Superior California' = c("Butte", "Colusa", "El Dorado", "Glenn", "Lassen", "Modoc", "Nevada", "Placer", "Plumas", "Sacramento", "Shasta", "Sierra", "Siskiyou", "Sutter", "Tehama", "Yolo", "Yuba"), 
                               'North Coast' = c("Del Norte", "Humboldt", "Lake", "Mendocino", "Napa", "Sonoma", "Trinity"), 
                               'San Francisco Bay Area' = c("Alameda", "Contra Costa", "Marin", "San Francisco", "San Mateo", "Santa Clara", "Solano"), 
                               'Northern San Joaquin Valley' = c("Alpine", "Amador", "Calaveras", "Madera", "Mariposa", "Merced", "Mono", "San Joaquin", "Stanislaus", "Tuolumne"), 
                               'Central Coast' = c("Monterey", "San Benito", "San Luis Obispo", "Santa Barbara", "Santa Cruz", "Ventura"), 
                               'Southern San Joaquin Valley' = c("Fresno", "Inyo", "Kern", "Kings", "Tulare"), 
                               'Inland Empire' = c("Riverside", "San Bernardino"), 
                               'Los Angeles County' = "Los Angeles", 
                               'Orange County' = "Orange", 
                               'San Diego - Imperial' = c("Imperial", "San Diego")))

ca_childcare
```


**4. Let's consider the median household income of each region, and how that income has changed over time. Create a table with ten rows, one for each region, and two columns, one for 2008 and one for 2018. The cells should contain the `median` of the median household income (expressed in 2018 dollars) of the `region` and the `study_year`. Arrange the rows by 2018 values.**

```{r}
#| label: median-income-by-region-over-time
# Hint: You will want to calculate the median of median income before you make columns for 2008/2018 by transforming the data

# Median household income code = `mhi_2018`
ca_median_income <- ca_childcare |> 
  select(region, study_year, mhi_2018) |> 
  filter(study_year %in% c(2008, 2018)) |> 
  group_by(region, study_year) |> 
  summarise(mean_mhi = mean(mhi_2018), .groups = "drop") |> 
  pivot_wider(names_from = study_year, values_from = mean_mhi)
  
ca_median_income
```


**5. Which California `region` had the lowest `median` full-time median weekly price for center-based childcare for infants in 2018? Does this `region` correspond to the `region` with the lowest `median` income in 2018 that you found in Q4?**  

*Hint: The code should give me the EXACT answer. This means having the code output the exact row(s) and variable(s) necessary for providing the solution. Consider the `slice()` functions.*

```{r}
#| label: lowest-median-weekly-price-2018

# Median weekly price for center-based childcare for infants = mc_infant
ca_mc_infants <- ca_childcare |> 
  select(region, study_year, mc_infant) |>
  filter(study_year == "2018") |>
  group_by(region, study_year) |> 
  summarise(mean_mc_infant = mean(mc_infant), .groups = "drop") |> 
  pivot_wider(names_from = study_year, values_from = mean_mc_infant) |> 
  slice_min(`2018`, n = 1)

ca_median_income |> 
  slice_min(`2018`, n = 1)

ca_mc_infants
```

> The California region that had the lowest median full-time median weekly price for center-based childcare for infants in 2018 was Superior California (241,7306). This region does not correspond to the region with the lowest median household income in 2018 (Southern San Joaquin Valley, 51599.4).


**6. Recreate the plot (look at online instructions) that shows, for all ten regions, the change over time of the full-time median price for center-based childcare for infants, toddlers, and preschoolers. Recreate the plot. You do not have to replicate the exact colors or theme, but your plot should have the same content, including the order of the facets and legend, reader-friendly labels, axes breaks, and a  smooth function.**


```{r, message = FALSE, fig.width=12, fig.height=5} 
#| label: recreate-plot

# Data transformation 
ca_mc_all <- ca_childcare |> 
  select(study_year, region, mc_infant, mc_toddler, mc_preschool) |> 
  pivot_longer(cols = starts_with("mc"), 
               names_to = "Age_Group", 
               values_to = "Median_Cost") |>
  mutate(Age_Group = factor(Age_Group, levels = c("mc_infant", 
                                                  "mc_toddler", 
                                                  "mc_preschool")))

# Graph
ggplot(ca_mc_all, aes(x = study_year, y = Median_Cost, color = factor(region))) + 
  geom_point() + 
  geom_smooth() + 
  facet_wrap(~ Age_Group, labeller = labeller(Age_Group = c(mc_infant = "Infant", 
                                                            mc_toddler = "Toddler", 
                                                            mc_preschool = "Preschool"))) + 
  labs(title = "Weekly Median Price for Center-Based Childcare ($)", 
       x = "Study Year", 
       y = NULL, 
       color = "California Region") + 
  scale_x_continuous(breaks = c(2008, 2010, 2012, 2014, 2016, 2018)) + 
  scale_y_continuous(limits = c(100, 500)) + 
  theme_bw() 

ca_mc_all
```


